Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-10 image_analysis_paper_development2/manuscript.tex
updated wing disc

Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
%
We also provide a software implementation of our methodology to aid in the application of our techniques to other experimental data sets. 
\vspace{1em}

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the difference between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that pairs of data point which are close are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parameterizing the curve.}
\label{fig:schematics}
\end{figure}

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
%
Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
%
We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
%
%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a movie of zebrafish epiboly. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
%
We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage.
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization).
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.
%
During this stage, the cells rearrange to form a uniform layer about four cells thick.
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole.
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A).
%
We then used vector diffusion maps to register and order the frames.
%
The results are shown in \fig~\ref{fig:zebrafish}B.
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C).
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.
%
Some errors in ordering images of the early embryo result from slow cell movement during the early developmental stage where cells divide and accumulate above the yolk.
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.

@@

We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo.
%
In practice, we are interested in cases where image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such variations,
we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability.
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.

@@

\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks. (A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack. Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray). Projections along the $x$-, $y$-, and $z$-axes are shown. (B) Example three-dimensional images, ordered using diffusion maps. The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427. (C) The average developmental trajectory for the registered and ordered images. }
\label{fig:wing_disc}
\end{figure*}

@@

Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computational prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
%
However, ordering data via diffusion maps is readily possible for registered three-dimensional data.
%
Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.

@@

For this particular system, the face of the imaginal disc is significantly larger than the edge of the disc, so that images will always be taken with the face towards the microscope.
%
Therefore, we need not consider registration in all three dimensions, as the z-axis will always be aligned with the short axis of the disc.
%
We can instead focus on registering the discs with respect to rotations only in the x-y plane.
%
We aligned the maximum intensity projections using angular synchronization, and then used these rotations to register the full three-dimensional in the x-y plane.
%
Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples.
%
Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
%
%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.

>>>>

Dynamics of developmental progress are commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
%
We also provide a software implementation of our methodology to aid in the application of our techniques to other experimental data sets.
\vspace{1em}

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

\begin{figure}[t]
\includegraphics{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. (A) Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. (B) The vectors from A, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in A: the difference between a pair of angles in B is the same as the pairwise angle in A. (C) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through a Gaussian kernel (see \SI), so that pairs of data point which are close are connected by darker (``stronger'') edges. (D) The data in C, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the perceived curve arclength, thus parametrizing the curve.}
\label{fig:schematics}
\end{figure}

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
%
Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
%
We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
%
%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a movie of zebrafish epiboly. Scale bar indicates 200$\mu$m. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
%
We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage.
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization).
%
In this experiment, embryos were immobilized for imaging so that the position and orientation remained fixed \cite{kane1996zebrafish}.
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.
%
During this stage, the cells rearrange to form a uniform layer about four cells thick.
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole.
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the two-dimensional frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A).
%
We then used vector diffusion maps to register and order the frames.
%
The results are shown in \fig~\ref{fig:zebrafish}B.
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C).
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.
%
Some errors in ordering images of the early embryo result from slow cell movement during the early developmental stage where cells divide and accumulate above the yolk.
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.

@@

We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo.
%
In practice, we are interested in cases where each image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such variations,
we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability.
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.

@@

\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks. (A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack. The anterior-posterior (A-P) and dorsal-ventral (D-V) axes are indicated. Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray). Projections along the $x$-, $y$-, and $z$-axes are shown. (B) Example three-dimensional images, ordered using diffusion maps. The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427. (C) The average developmental trajectory for the registered and ordered images. }
\label{fig:wing_disc}
\end{figure*}

@@

Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computationally prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
%
However, ordering data via diffusion maps is readily possible for pre-registered three-dimensional data.
%
Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.

@@

In the wing disc, the anterior-posterior and dorsal-ventral axes are significantly larger than the third principal axis (see \fig~\ref{fig:wing_disc}A). 
%
Therefore, we need not consider registration in all three dimensions, as the z-axis for imaging (the axis perpendicular to the imaging plane) will always be aligned with the short axis of the wing disc.
%
We can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
%
To register the data, we first aligned the maximum intensity projections using angular synchronization.
%
We then used these rotations to register the full three-dimensional data in the x-y plane.
%
Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples.
%
Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
%
%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.
